### 2016-05-07 Projects/DelCo_2012/Report/DelCo_2012.tex
Added Refs and revised Intro and sec 2
Added Cal Poly refs and rewords section 1.  Moved nozzle info to Sec 2.1
and added a 2.0 paragraph

A number of fire departments throughout the US have acquired CAFS units for structural firefighting, but have curtailed the use of these units for this purpose until potential safety implications are better understood. This project is expected to produce a better understanding of the effectiveness and safety implication of CAFS for structural firefighting through the various tasks associated with this project, including: 1) A literature review of available published information on the use of CAFS for structural firefighting~\cite{Mitchell:1}; 2) A workshop that brings interested and experienced parties together to discuss safety implications associated with this project and to provide guidance on the proposed experimental research plan~\cite{Grant:2011}; 3) The performance and analysis of a number of well-instrumented large-scale enclosure fire tests using both CAFS and plain water to evaluate the effectiveness of CAFS for interior structural firefighting; and 4) The performance and analysis of a number of fire ground evolutions to evaluate differences between CAFS and plain water hose lines related to issues such as effective hose stream throw and distribution, forces needed to carry and move hoses streams both horizontally and vertically, and forces need to kink hose lines. The conclusions should include much better information regarding the safe and effective use of CAFS for structural firefighting.

@@

Compressed air foam (CAF) generated with a solution of water and a Class A foam concentrate has been in use by wildland fire agencies since the late 1970’s. The wildland fire community found that the Class A foam reduced the surface tension of water, which improved water's ability to penetrate and soak into wildland fuels. A CAFS could turn a Class A foam solution into a blanket of bubbles that would allow the wildland firefighters to coat the fuels. The additional ``soak'' time provided by the foam layer allows the water to penetrate the hot surface of a burning fuel, and also provides additional time for water to penetrate into fuels being pre-treated in an effort to prevent ignition. An added feature of the blanket of bubbles is that firefighters could see where they had applied the agent so that they could make a more effective application. In the wildland, water for firefighting is typically a precious resource, is in limited supply, and has to be carried to where it is needed. CAF increases the effectiveness of water, and may reduce the quantity required to suppress fires and prevent the ignition of adjacent fuels. Last but not least the compressed air adds energy to the water stream providing additional reach and volume to a hose stream with a relatively low mass flow rate of water. These CAFS advantages combine to provide wildland firefighters with a tool that could be used to pre-treat homes and structures to protect them from an approaching fire, or could be used for an exterior (indirect) fire suppression attack on a structure.

In the decades that followed the development of CAFS, a number of urban/suburban fire departments across the country began to integrate CAFS into their apparatus. For example, the Los Angeles County Fire Department incorporated CAFS into their apparatus for wildland urban interface fires, but also examined using them for structural fire fighting. Departments across the country began to experiment with CAFS systems for use in structural firefighting. In most documented cases, the findings about the effectiveness of the CAFS were favorable, yet after decades of fire department trials, CAFS still has not been significantly incorporated into structural firefighting tactics. The importance of tactics is illustrated by the example of one community, which after putting a fleet of CAFS equipped fire engines into service, has now issued an order that the CAFS should not be used after incidents occurred where firefighters were burned while performing interior suppression operations in a structure.

There has never been a comprehensive scientific study on the use of CAFS for structural fire fighting.  Fire departments have conducted pilot programs to examine implementation and effectiveness of CAFS, foam suppliers and equipment manufacturers have conducted minimally instrumented experiments, typically in acquired structures, and test laboratories and universities have examined small sections of the CAFS puzzle, but none of these efforts has resulted in the body of knowledge needed by the fire service~\cite{Madrzykowski:3,USFA:CAFS,Colletti:1,Boyd:1,Tafreshi:1,Madrzykowski:4,Colletti:2,Tafreshi:2,Lorh:2002,Colletti:3,Kim:2012}. Much of the data has been generated by people and organizations involved in selling the technology to the fire service and, in many cases, the conclusions are based on a very limited set of instruments or comparisons between water and form that were not designed as scientific experiments.

@@

Based on the results of the NIST 1998 study, the fire fighting agents, commonly referred to as Class A foams, were superior to water in reducing contact angle, cooling and penetrating the fuels, mass retention on the fuels and ignition inhibition of the fuels.  When applied in the form of foam, the water agent solutions provide enhanced radiation exposure protection to a structure relative to plain water.  The foams enhance the water by keeping the water in place on vertical surfaces such as walls and the reduced inter-facial surface tension allows the water to penetrate absorbent fuels.  These enhancements also make the agents valuable tools for mop-up. The other experiments conducted during this study: specific heat, drop size, wood crib suppression, smoke generation and Class D fire suppression yielded similar results for both Class A foams and plain water.

@@

Since Montgomery County (MD) Fire and Rescue Service (MCFRS) provided the CAFS equipped engines and the staffing needed to provide both the water and CAF for the experiments. Baseline operating conditions were agreed upon at the workshop for these experiments (Table~\ref{tab:op_condition}). These experiments were conducted at the Delaware County Emergency Services Training Center (ESTC) in Sharon Hill, PA.

@@

\section{Description of Experiments}
\label{sec:desc_experiments}
The experimental setup for this test series is described by the three types of experiments conducted: spray density measures, gas layer cooling, and fire suppression. The nozzles chosen for these experiments were decided upon during the CAFS workshop~\cite{Grant:2011}: Task Force Tips (TFT) Metro 1, fixed gallonage nozzles with a flow rate of 120 gpm. These are automatic nozzles which have a 7/8~inch slug-tip and can be varied from a straight stream to a wide fog for application. The set-up also included 100~ft of 1-3/4~inch hose. Nozzles were mounted to a ``blitz-fire'' monitor device and placed inside the structures. The nozzles were controlled and operated via a gate valve from the exterior.

>>>>

A number of fire departments throughout the US have acquired CAFS units for structural firefighting, but many have limited or stopped the use of these CAF for structural firefighting due to operational challenges or due to lack of understanding related to fireground safety issues. This project is expected to produce a better understanding of the effectiveness and safety implications of CAF for structural firefighting through the various tasks associated with this project, including: 1) A literature review of available published information on the use of CAFS for structural firefighting~\cite{Mitchell:1}; 2) A workshop that brings interested and experienced parties together to discuss safety implications associated with this project and to provide guidance on the proposed experimental research plan~\cite{Grant:2011}; 3) The performance and analysis of a number of well-instrumented large-scale enclosure fire tests using both CAFS and plain water to evaluate the effectiveness of CAFS for interior structural firefighting; and 4) The performance and analysis of a number of fire ground evolutions to evaluate differences between CAFS and plain water hose lines related to issues such as effective hose stream throw and distribution, nozzle reaction force, the force to kink the hose, potential for slipping on a variety of surfaces and flow separation~\cite{Carracino:2013,Dicus:2012,LaPolla:2012}. The research findings are provide information with the potential to improve the safe and effective use of CAFS for structural firefighting.

@@

Compressed air foam (CAF) generated with a solution of water and a Class A foam concentrate has been in use by wildland fire agencies since the late 1970's. The wildland fire community found that the Class A foam reduced the surface tension of water, which improved water's ability to penetrate and soak into wildland fuels. A CAFS could turn a Class A foam solution into a blanket of bubbles that would allow the wildland firefighters to coat the fuels. The additional ``soak'' time provided by the foam layer allows the water to penetrate the hot surface of a burning fuel, and also provides additional time for water to penetrate into fuels being pre-treated in an effort to prevent ignition. An added feature of the blanket of bubbles is that firefighters could see where they had applied the agent so that they could make a more effective application. In the wildland, water for firefighting is typically a precious resource, is in limited supply, and has to be carried to where it is needed. CAF increases the effectiveness of water, and may reduce the quantity required to suppress fires and prevent the ignition of adjacent fuels. Last but not least the compressed air adds energy to the water stream providing additional reach and volume to a hose stream with a relatively low mass flow rate of water. These CAFS advantages combine to provide wildland firefighters with a tool that could be used to pre-treat homes and structures to protect them from an approaching fire, or could be used for an exterior (indirect) fire suppression attack on a structure.

In the decades that followed the development of CAFS, a number of urban/suburban fire departments across the country began to integrate CAFS into their apparatus. For example, the Los Angeles County Fire Department incorporated CAFS into their apparatus for wildland urban interface fires, but also examined using them for structural fire fighting. Departments across the country began to experiment with CAFS systems for use in structural firefighting. In most documented cases, the findings about the effectiveness of the CAF were favorable, yet after decades of fire department trials, CAF has still not been significantly incorporated into structural firefighting tactics. This is illustrated by the example of one community, which after putting a fleet of CAFS equipped fire engines into service, had issued an order that the CAFS should not be used after incidents occurred where firefighters were burned while performing interior suppression operations with CAF inside a structure.

There has never been a comprehensive scientific study on the use of CAF for structural fire fighting.  Fire departments have conducted pilot programs to examine implementation and effectiveness of CAF, foam suppliers and equipment manufacturers have conducted minimally instrumented experiments, typically in acquired structures, and test laboratories and universities have examined small sections of the CAF puzzle, but none of these efforts has resulted in the body of knowledge needed by the fire service~\cite{Madrzykowski:3,USFA:CAFS,Colletti:1,Boyd:1,Tafreshi:1,Madrzykowski:4,Colletti:2,Tafreshi:2,Lorh:2002,Colletti:3,Kim:2012}.

@@

Based on the results of the NIST 1998 study, the fire fighting agents, commonly referred to as Class A foams, were superior to water in reducing contact angle, cooling and penetrating wood based fuels, mass retention on the wood fuels and ignition inhibition of the wood fuels.  When applied in the form of foam, the water agent solutions provide enhanced radiation exposure protection to a structure relative to plain water.  The foams enhance the water by keeping the water in place on vertical surfaces such as walls and the reduced inter-facial surface tension allows the water to penetrate absorbent fuels.  These enhancements also make the agents valuable tools for mop-up. The other experiments conducted during this study: specific heat, drop size, wood crib suppression, smoke generation and Class D fire suppression yielded similar results for both Class A foams and plain water.

@@

All of the experiments were conducted at full scale in compartments and structures that were representitive room sizes in a residential occupancy.  Technical panel members whose departments rountinely use CAF for structural firefighting were invited and present at many of the fire suppression experiments.  The series of experiments, planned during the workshop, for examining spray density measures, gas layer cooling, and fire suppression were conducted in the fall of 2012.  The data from these experiments were presented to the technical panel member.   The fire suppression experiments were conducted with a fuel package that included upholstered furniture.  After a review of the data, a decision was made to conduct an additional series of suppression experiments with a fuel package was composed mainly of wood based materials, as a way to increase the challenge of the suppression experiment.  After the wood based fuel package experiments were completed, it was decided in discussions with members of the technical panel that other experimental conditions needed to be examined: room size, additional ventilation, and a manually operated handline.  The issue of a larger fire compartment had also been raised durning the workshop discussion, but the desire for replicate experiments and financial limitations removed the large floor plan experiments from the original plan.  In order to accomplish the last series of experiments, additional funding was obtained from NIST and the experiments were conducted in 2015.

The sections below describe the types of experiments, the experimental facilities, instrumentation, and fuel loads.  Project partner, Montgomery County (MD) Fire and Rescue Service (MCFRS) provided the CAFS equipped engines and the staffing needed to provide both the water and CAF for the experiments. These experiments were conducted at the Delaware County Emergency Services Training Center (ESTC) in Sharon Hill, PA.

\section{Description of Experiments}
\label{sec:desc_experiments}
The experimental setup for this test series is described by the three types of experiments conducted: spray density measures, gas layer cooling, and fire suppression.   Baseline operating conditions were agreed upon at the workshop for these experiments (Table~\ref{tab:op_condition}).   Baseline operating conditions were agreed upon at the CAFS workshop for these experiments (Table~\ref{tab:op_condition})~\cite{Grant:2011}.   The nozzles chosen for these experiments were: Task Force Tips (TFT) Metro 1, fixed gallonage nozzles with a flow rate of 120 gpm. These are automatic nozzles which have a 7/8~inch slug-tip and can be varied from a straight stream to a wide fog for application. The set-up also included 100~ft of 1-3/4~inch hose. Nozzles were mounted to a ``blitz-fire'' monitor device and placed inside the structures. The nozzles were controlled and operated via a gate valve from the exterior.

@@
